Reproducible Research - Peer Assessment 1
=========================================

# Synopsis
This report is a submission for the Peer Assessment #1 for the Coursera Data Science specialization, Reproducible Research course.  In this assignment, we look at the data from an activity measuring device used by a particular individual over a period of two months. This particular device measures the number of steps taken within 5 minute intervals throughout the day.

# Loading and Processing the Raw Data

The data for this assignment is available on the Coursera course site  [Activity monitoring data (52k)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip), along with the [instructions and additional background](https://class.coursera.org/repdata-002/human_grading/view/courses/972084/assessments/3/submissions) from which the following explanation is excerpted.

[Excerpt from assignment instructions]

The variables included in this dataset are:
- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


## Reading in the data
```{r load-data}
rawdata <- read.csv('project/data/activity.csv')
rawdata$date = as.Date(rawdata$date)

head(rawdata)
```

# Assignment questions

## Q1: What is mean total number of steps taken per day?

For this question, missing values are ignored (as per assignment instructions).

```{r}
library(plyr)
byDay <- ddply(rawdata, .(date), summarize,
               sum=sum(steps,na.rm=T),mean=mean(steps,na.rm=T),
               median=median(steps,na.rm=T), count=length(steps),
               nonZero=sum(!is.na(steps) & steps != 0), zero=sum(!is.na(steps) & steps == 0),
               na=sum(is.na(steps)))

mean(byDay$sum)
median(byDay$sum)

```

```{r fig.width=7, fig.height=6}
hist(byDay$sum,
    main="Distribution of the number of steps per day",
    xlab="Number of steps"
     )

```


## Q2: What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
byInterval <- ddply(rawdata, .(interval), summarize,
                    sum=sum(steps,na.rm=T),mean=mean(steps,na.rm=T), median=median(steps,na.rm=T),
                    count=length(steps), nonZero=sum(!is.na(steps) & steps != 0),
                    zero=sum(steps == 0), na=sum(is.na(steps)))
tMax <- byInterval[which.max(byInterval$mean),1]

tMax
```

Here is a plot showing the average steps across days, with a vertical line marking the maximum.

```{r fig.width=7, fig.height=6}
plot(byInterval$interval, byInterval$mean, type="l",
    main="Mean number of steps by interval, averaged across days",
    xlab="interval",
    ylab="Mean number of steps"
)
lines(c(tMax,tMax),c(0,max(byInterval$mean)), type="l", col="red")
text(tMax, max(byInterval$mean),
     labels=c(
       sprintf("max (%d, %.1f)", tMax, round(max(byInterval$mean)),1)
     )
)

```


## Q3: Imputing missing values



```{r}
# missing values
sum(is.na(rawdata$steps))
```


By looking at the data, we see that most days have no missing data, and the rest have no data.
```{r fig.width=7, fig.height=6}
with(byDay,plot(date, na, ylab="Number of missing data points"), main="Missing data points per day")

```

Without other knowledge, a reasonable estimation for each missing interval's data is the mean for that interval.
```{r}
processedData <- data.frame(rawdata)

for(i  in 1:nrow(byInterval)) {
  processedData$steps[is.na(processedData$steps) & processedData$interval == byInterval$interval[i]] = byInterval$mean[i]
}

```

After the imputation, the mean and median are higher, and equal to one another. The histogram shows a smoother distribution.

```{r}
byDay2 <- ddply(processedData, .(date), summarize,
               sum=sum(steps,na.rm=T),mean=mean(steps,na.rm=T),
               median=median(steps,na.rm=T), count=length(steps),
               nonZero=sum(steps != 0), zero=sum(steps == 0),
               na=sum(is.na(steps)))

mean(byDay2$sum)
median(byDay2$sum)

```


```{r fig.width=7, fig.height=6}
hist(byDay2$sum,
    main="Distribution of the number of steps per day (with imputed data)",
    xlab="Number of steps"
     )

```


## Q4: Are there differences in activity patterns between weekdays and weekends?


```{r}
processedData$daytype = as.factor( ifelse(weekdays(processedData$date) %in% c("Saturday","Sunday"), "weekend","weekday"))
byInterval2 <- ddply(processedData, .(daytype,interval), summarize,
                    sum=sum(steps), mean=mean(steps), median=median(steps)
                    )

```

By plotting the mean steps per interval separately for weekends and weekdays, it appears that the number of steps is spread more evenly on the weekends.

```{r fig.width=7, fig.height=6}
library(lattice)
xyplot(mean ~ interval | daytype, data=byInterval2, type="l", layout=c(1,2),
    main="Mean steps by interval, averaged across days (with imputed data)",
    xlab="interval",
    ylab="Mean steps"
   )
```

